Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-19 articles/cycle/response-referee1.tex

% ----------------------------------------------------------------------

\sechead{Specific comments}

    \referee{\textbf{p.~4, l.~9--10:}
        Can the physical basis of englacial water fraction and sensitivity of
        results be summarized briefly? This is not a usual component in
        ice-sheet models. Is the cap value ("capped at 0.01") well constrained,
        and does it have a significant effect on results?}

    Laboratory experiments have demonstrated that the rheology of temperate,
    polycrystalline ice depends on its content in liquid water
    \citep[p.~65--66]{Cuffey.Paterson.2010}. Unfortunately, the only
    measurements available to date \citep{Duval.1977}, used to quantify the
    effect of liquid water on ice softness, the creep parameter $A$ in Glen's
    flow law \citep{Lliboutry.Duval.1985}, only extend to fractions of liquid
    water content between 0 and 0.8\%. They show a three-fold increase of ice
    softness over this range \citep[Fig.~1]{Duval.1977}.

    Although ice sheet models have previously ignored this effect, it has now
    become a typical component of polythermal models such as SICOPOLIS
    \citep{Greve.1997}, COMICE \citep{Ruckamp.etal.2010}, PISM
    \citep{Aschwanden.etal.2012}, ISSM \citep{Seroussi.etal.2013}, and
    TIM-FD$\unit{^{3}}$ \citep{Kleiner.Humbert.2014}.

    Water fractions between 1 and 5\,\% have repreatedly been observed in
    temperate glaciers \citep{Murray.etal.2000, Murray.etal.2007,
    Bradford.Harper.2005, Bradford.etal.2009}, and also occur in model results
    \citep[e.g.,][]{Blatter.Greve.2015}, but it is not known whether ice
    viscosity continues to decrease substancially for values above 0.8\,\%.
    Previous modelling studies have commonly assumed constant ice viscosity
    above 1\,\%. This arbitrary threshold is not constrained at all, and the
    urgent need for new ice deformation experiments has been pointed out
    \citep{Kleiner.etal.2015}.

    The sensitivity of our results to the 1\,\% threshold was not tested.
    However, in our model results, liquid water fractions above 1\,\% typically
    only occur within the basal temperate layer of the fastest-moving glaciers
    where ice movement is dominated by basal sliding. We suspect that a local
    increase of ice deformation there is negligible in comparaison to
    uncertainties caused by basal sliding and more importantly climate forcing.

    We prefer to avoid including the above discussion in the manuscript, but we
    have reworked the sentence on water content:

    \msquote{%
        [Ice softness] increases with liquid water fractions up to 0.01
        \citep[p.~65--66]{Duval.1977, Lliboutry.Duval.1985,
        Cuffey.Paterson.2010}, an arbitrary threshold above which new
        measurements are critically needed \citep{Kleiner.etal.2015}.}

    The uncertainty to unknown rheology of water-rich temperate ice was also
    mentioned in the conclusions.

@@

        Done.

    \referee{\textbf{Fig.~1 caption, l.~4:}
        Perhaps change ``estimated'' to ``estimate'', or ``estimated of'' to
        ``estimated''.}

        Done.

\sechead{References}

\bibliographystyle{copernicus}
\bibliography{../../references/references}

### 2018-07-19 articles/cycle/alps-cycle.tex
Assess referee 1 comment on water content.

    Ice deformation is governed by the constitutive law for ice
    \citep{Glen.1952, Nye.1953}. Ice softness depends on ice temperature,
    pressure, and water content, through an enthalpy scheme
    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012} and a piece-wise
    Arrhenius-type rheology representative of Holocene polar ice
    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}.
    The englacial water fraction is
    capped at a~maximum value of 0.01, above which no measurements are
    available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}.

@@

    However, these results are limited by the simplicity of the climate
    forcing and surface mass balance model applied, and uncertainties in ice
    flow physics. In particular, additional palaeo-climate variability over
    the European Alps may have caused more glaciations than modelled here.
    Using more specifically targeted sensitivity runs, future
    modelling studies will certainly be able to quantify uncertainties
    associated with some of the above limitations. Nevertheless, we hope that
    these conclusions will also serve as a basis for future studies of glacial
    geology in the Alps, and call for a more systematic aggregation and
    homogenization of glacial geological data to form a basis for model
    validation across the entire Alpine range.

>>>>

    Ice deformation is governed by the constitutive law for ice
    \citep{Glen.1952, Nye.1953}. Ice softness depends on ice temperature,
    pressure, and water content, through an enthalpy scheme
    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012}. It follows a
    piece-wise Arrhenius-type rheology representative of Holocene polar ice
    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}, and increases
    with liquid water fractions up to 0.01 \citep[p.~65--66]{Duval.1977,
    Lliboutry.Duval.1985, Cuffey.Paterson.2010}, an arbitrary threshold above
    which new measurements are critically needed \citep{Kleiner.etal.2015}.

@@

    However, these results are limited by uncertainties on basal sliding and
    on the unknown rheology of water-rich temperate ice. The potential
    variability in surface mass balance parameters, as well as the simplicity
    of the climate forcing, are even more important.
    In particular, additional palaeo-climate variability over
    the European Alps may have caused more glaciations than modelled here.
    Using more specifically targeted sensitivity runs, future
    modelling studies will certainly be able to quantify uncertainties
    associated with some of the above limitations. Nevertheless, we hope that
    these conclusions will also serve as a basis for future studies of glacial
    geology in the Alps, and call for a more systematic aggregation and
    homogenization of glacial geological data to form a basis for model
    validation across the entire Alpine range.
